---
title: "MovieLens  \nCreating a movie recommendation system"
author: "Anton Ivanov"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Introduction

In this project, we will be creating a movie recommendation system using the [10M version of the MovieLens dataset](https://grouplens.org/datasets/movielens/10m/).
This is a small subset of a larger dataset containing 10 million ratings for 10,000 movies from 72,000 users.

Recommendation systems are of particular importance in the modern world. They allow companies such as online shops or streaming services to recommend users matching goods or content, providing higher profit.

The goal of the project is to predict a rating for a specific movie from a specific user, based on the available ratings.
The used recommendation system is as following: One star suggests it is not a good movie, whereas five stars suggests it is an excellent movie.

```{r complete dataset, message=FALSE}
library(tidyverse)
load("movielens_s.RData")
as_tibble(movielens_s)
```

The data will be split into
The dataset includes ratings for  movies from users.
